Import the necessary libraries

```{r}
library(usethis)
library(here)
library(sf)
library(tmap)
library(raster)
library(sp)
library(gstat)
library(geoR)
```

Load in quantitative datasets

```{r}
SO2_Emissions <- read.csv(here("wk6_dataset", "US 2019 SO2 Emissions data.csv"))

# inspect
head(SO2_Emissions, 10)
```

Load in the shapefiles

```{r}
US_Nation_Border <- st_read(here("wk6_dataset", "US Nation Border.shp"))
US_State_Border <- st_read(here("wk6_dataset", "US State Borders.shp"))

# inspect
qtm(US_Nation_Border)
qtm(US_State_Border)
```

Data Preparation

```{r}
# coerce the spreadsheet into an sf object
# First tell R that it's coordinates are currently in decimal degrees (i.e. WGS84) before the transformation
SO2_Emissions_sf <- st_as_sf(SO2_Emissions, coords = c("Longitude", "Latitude"), crs = 4326)
# now apply the transformation from WGS84 to Mercator i.e. crs = 3857
SO2_Emissions_sf <- st_transform(SO2_Emissions_sf, 3857)
# inspect CRS details
st_crs(SO2_Emissions_sf)
```

Map out the points quickly

```{r}
tm_shape(US_Nation_Border) +
  tm_polygons(alpha=0, border.col="black") +
tm_shape(SO2_Emissions_sf) +
  tm_dots() +
tm_scale_bar(position = c("left", "bottom")) +
tm_compass(position = c("right", "bottom"))
```

Plotting the Empirical Semivariogram

Use the function `r variogram()` to create the object for plotting the empirical variogram

```{r}
# coerce SO2_Emissions_sf to be a 'sp' spatial dataframe object as it's variogram does not use 'sf' objects

# ignore warning message
SO2_Emissions_sf <- as(SO2_Emissions_sf, "Spatial")
# use variogram() function to compute the semivariance with a null model Mean_SO2 as outcome

SO2_emp.variogram <- variogram(Mean_SO2~1, SO2_Emissions_sf)
# compute the object to reveal a table
SO2_emp.variogram
# plot
plot(SO2_emp.variogram)
```

Plotting the Theoretical Semivariogram

```{r}
# fit exponential
exp_SO2_emp.variogram <- fit.variogram(SO2_emp.variogram, model = vgm(65, "Exp", 1180000, 17))

exp_SO2_emp.variogram
plot(SO2_emp.variogram, exp_SO2_emp.variogram, main = "Exponential Model (Nug: 3.6, PSill: 55.9, Range: 296255m")
```

```{r}
# fit spherical
sph_SO2_emp.variogram <- fit.variogram(SO2_emp.variogram, model = vgm(65, "Sph", 1180000, 17))
sph_SO2_emp.variogram
plot(SO2_emp.variogram, sph_SO2_emp.variogram, main = "Spherical Model (Nug: 10.5, PSill: 49.6, Range: 857452m")
```

```{r}
# fit gaussian
gau_SO2_emp.variogram <- fit.variogram(SO2_emp.variogram, model = vgm(65, "Gau", 1180000, 17))
gau_SO2_emp.variogram
plot(SO2_emp.variogram, gau_SO2_emp.variogram, main = "Gaussian Model (Nug: 12.8, PSill: 39.1, Range: 244807m")
```

Which one is the best model?

```{r}
# select the best model
best_SO2_emp.variogram <- fit.variogram(SO2_emp.variogram, model = vgm(c("Exp", "Gau", "Sph")))
best_SO2_emp.variogram
```

Kriging Modelling

Build a blank raster templte to begin with:

```{r}
RasterTemplate <- raster(SO2_Emissions_sf)
res(RasterTemplate) <- 5000
# declare template as a spatial grid
grid.interpolation <- as(RasterTemplate, 'SpatialGrid')
```

Implement spatial interpolation on blank template using Kriging

```{r}
modelKrigingExp <- gstat(formula = Mean_SO2~1, locations = SO2_Emissions_sf, model = exp_SO2_emp.variogram)
```

```{r}
# this may take roughly 5 minutes
Kriged_SO2 <- predict(modelKrigingExp, grid.interpolation)
```
The above analysis produces two separate rasters: i) Predicted SO2 and ii) Variation in SO2

Exporting the results as .tiff format thematics visualisation in `r tmap`


```{r}
# save both the prediction and variance as a multi-layer stack (a 'brick' raster):
brickedKriged_SO2_Results <- brick(Kriged_SO2)
```

We can save them individually from the multi-layer raster as follows:

```{r}
# Separate the rasters accordingly
exp.prediction <- raster(brickedKriged_SO2_Results, layer = 1)
exp.variance <- raster(brickedKriged_SO2_Results, layer = 2)
# save the output locally on your computer
writeRaster(exp.prediction, "Predicted SO2 levels in USA.tif", format="GTiff", overwrite = TRUE)
writeRaster(exp.variance, "Variance SO2 levels in USA.tif", format="GTiff", overwrite = TRUE)
```

Before visualising the outputs, we first need to make the values of the raster predictions made outside of US Border. For this, use the `r mask()` function

```{r}
US_Nation_Border_sp_shp <- as(US_Nation_Border, "Spatial")
exp.prediction_masked <- mask(exp.prediction, US_Nation_Border)
```

Use `r st_bbox()` function on the `r US_Nation_Border` shape file object to get the extent of the region

```{r}
frameExtent <- st_bbox(US_Nation_Border)
frameExtent
```

Visualise the predictions:

```{r}
tm_shape(exp.prediction_masked, bbox = frameExtent) +
  tm_raster(title = "Predicted SO2 ppb", style = "cont", palette ="Reds") +
tm_shape(US_State_Border) +
  tm_polygons(alpha = 0, border.col = "black") +
  tm_text("STUSPS", size = "AREA") +
tm_shape(SO2_Emissions_sf) +
  tm_dots() +
tm_scale_bar(position = c("left", "bottom")) +
tm_compass(position = c("right", "bottom")) +
tm_layout(frame = FALSE, legend.title.size = 0.5, legend.text.size = 0.5, legend.position = c("right", "bottom"))
```

Reclassify the continuous values stored in the grid/pixels into discrete values using the following scheme

```{r}
# create a vector for the reclassification
reclassifyRaster <- c(0, 1, 0,
                      1, 5, 1,
                      5, 10, 2,
                      10, 15, 3,
                      15, 20, 4,
                      20, 30, 5,
                      30, 40, 6,
                      40, 70, 7)

# then store the values into a matrix
reclassifyRaster_Mat <- matrix(reclassifyRaster, ncol=3, byrow=TRUE)
reclassifyRaster_Mat
```

Apply the matrix to the raster object to reclassify the pixels

```{r}
exp.prediction_masked_recl <- reclassify(exp.prediction_masked, reclassifyRaster_Mat)
```

Visualise the predictions

```{r}
tm_shape(exp.prediction_masked_recl, bbox = frameExtent) +
  tm_raster(title = "Predicted SO2 ppb", style = "cat", palette = "Reds",
            labels = c("<1.0 ppb", "1.0-4.9 ppbs",
                       "5.0-9.9 ppbs", "10.0-14.9 ppbs",
                       "15.0-19.9 ppbs", "20.0-29.9 ppbs",
                       "30.0-39.9 ppbs", "+40.0 ppbs")) +
tm_shape(US_State_Border) +
  tm_polygons(alpha = 0, border.col = "black") +
tm_text("STUSPS", size = "AREA") +
tm_shape(SO2_Emissions_sf) +
  tm_dots() +
tm_scale_bar(position = c("left", "bottom")) +
tm_compass(position = c("right", "bottom")) +
tm_layout(frame = FALSE, legend.title.size = 0.5, legend.text.size = 0.5, legend.position = c("right", "bottom"))
```

You can also visualises the variance

```{r}
exp.variance_masked <- mask(exp.variance, US_Nation_Border)

tm_shape(exp.variance_masked, bbox = frameExtent) +
  tm_raster(title = "Variance SO2 ppb", style = "cont", palette = "Oranges") +
tm_shape(US_State_Border) +
  tm_polygons(alpha = 0, border.col = "black") +
tm_text("STUSPS", size = "AREA") +
tm_shape(SO2_Emissions_sf) +
  tm_dots() +
tm_scale_bar(position = c("left", "bottom")) +
tm_compass(position = c("right", "bottom")) +
tm_layout(frame = FALSE, legend.title.size = 0.5, legend.text.size = 0.5, legend.position = c("right", "bottom"))
```

Again, it will be better to reclassify the raster so we can es where the model predicts the SO2 with high and low errors

```{r}
reclassifyRaster_var <- c(0, 5, 0,
                      5, 10, 1,
                      10, 20, 2,
                      20, 30, 3,
                      30, 40, 4,
                      40, 50, 5,
                      50, 60, 6)

reclassifyRaster_Mat_var <- matrix(reclassifyRaster_var, ncol=3, byrow=TRUE)
reclassifyRaster_Mat_var
```

```{r}
exp.variance_masked_recl <- reclassify(exp.variance_masked, reclassifyRaster_Mat_var)

tm_shape(exp.variance_masked_recl, bbox = frameExtent) +
  tm_raster(title = "Variance SO2 ppb", style = "cat", palette = "Oranges",
            labels = c("<5.0 ppb", "5.0-9.9 ppbs",
                       "10.0-19.9 ppbs", "20.0-29.9 ppbs",
                       "30.0-39.9 ppbs", "40.0-49.9 ppbs",
                       "+50.0 ppbs")) +
tm_shape(US_State_Border) +
  tm_polygons(alpha = 0, border.col = "black") +
tm_text("STUSPS", size = "AREA") +
tm_shape(SO2_Emissions_sf) +
  tm_dots() +
tm_scale_bar(position = c("left", "bottom")) +
tm_compass(position = c("right", "bottom")) +
tm_layout(frame = FALSE, legend.title.size = 0.5, legend.text.size = 0.5, legend.position = c("right", "bottom"))
```